---
title: "APC Identification Problem: Simuatlion Experiment based on real data of 16 narwhals"
author: "Lars Nørtoft Reiter"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,fig.width = 8, fig.height = 6)

```

```{r}
library(ggplot2)
library(gridExtra)
library(reshape2)
library(dplyr)
library(tidyr)
library(mgcv)

df <- read.csv("narwhal_age_year_data.csv")
df$tusk <- as.factor(df$tusk)
```


### SIMULATION EXPERIMENT 1: Linear DECREASE in environmental concentration

```{r}
# Make calendar effect (linear)
calendar_period <- min(df$year):max(df$year)
sigma = 0.5
calendar_effect <- 205  - 0.1 * calendar_period  + rnorm(length(calendar_period), sd=sigma)

df_year <- data.frame(year = calendar_period, year_conc = calendar_effect)

# merge
df$year_conc <- NULL
df <- merge(df,df_year, by="year")

# plot
plot(calendar_period, calendar_effect, type="l")
abline(v=aggregate((df$year), list(df$tusk), min)$x, lty ="dashed", col="red")

# Make individual effect
df_age <- df %>% select(tusk = tusk, age = age, age_conc = age_conc)

# create individual intrinsic component which interacts with environmental concentraiton

# FIRST CHOICE: Exponential decay
# age_year_interaction <- function(x, year_conc){
#   # Exponential individual decay where baseline-factor interacts with yearly concentration in water + slight gamma noise
#   rnorm(length(year_conc), year_conc, 1)*exp(-rnorm(1,0.1,0.01)*x)+rgamma(length(x), 0.1, 1)
# }

# SECOND CHOICE: Constant level
age_year_interaction <- function(x, year_conc){
  # Constant individual level which interacts with yearly concentration in water
  year_conc + rgamma(length(year_conc), year_conc, 1)
}

# # EXAMPLE
# whale <- "888"
# tusk_df <- df_age[df_age$tusk==whale,]
# byIdx <- which(calendar_period %in% df[df$tusk==whale,]$year) # calendar period of whale
# baseline <- calendar_effect[byIdx]
# plot(tusk_df$age, age_year_interaction(tusk_df$age, baseline), type="l")

# create age-effect (confounded by year)
for(tusk in unique(df$tusk)){
  tusk_df <- df_age[df_age$tusk==tusk,]
  byIdx <- which(calendar_period %in% df[df$tusk==tusk,]$year) # calendar period of whale
  baseline <- calendar_effect[byIdx]
  df[df$tusk==tusk,]$age_conc <- age_year_interaction(tusk_df$age, baseline)
}

# plot
ggplot(df, aes(x=age, y=age_conc))+
  geom_line()+
  facet_wrap(.~tusk)

# Fit the model for this element
model <- gamm(age_conc ~ s(age, k=5, bs="cs"), 
              random = list(tusk = ~1), 
              data = df)

# Create prediction data
age_range <- seq(min(df$age, na.rm = TRUE), 
                 max(df$age, na.rm = TRUE), 
                 length.out = 100)

pred_data <- data.frame(age = age_range)

# Generate predictions with confidence intervals
predictions <- predict(model$gam, newdata = pred_data, se.fit = TRUE)

pred_data$fit <- predictions$fit
pred_data$se <- predictions$se.fit
pred_data$lower <- pred_data$fit - 1.96 * pred_data$se
pred_data$upper <- pred_data$fit + 1.96 * pred_data$se

# Create the plot
ggplot(df, aes(x = age, y = age_conc)) +
  geom_line(aes(color = tusk), alpha = 0.5) +
  geom_ribbon(data = pred_data, 
              aes(x = age, y = fit, ymin = lower, ymax = upper),
              alpha = 0.3, fill = "blue", inherit.aes = FALSE) +
  geom_line(data = pred_data, 
            aes(x = age, y = fit), 
            color = "blue", size = 1.2, inherit.aes = FALSE) +
  ylab("Annual concentration") +
  theme_minimal()
```



### SIMULATION EXPERIMENT 2: Linear INCREASE in environmental concentration


```{r}
# Make calendar effect (linear)
calendar_period <- min(df$year):max(df$year)
sigma = 0.5
calendar_effect <- -190  + 0.1 * calendar_period  + rnorm(length(calendar_period), sd=sigma)

df_year <- data.frame(year = calendar_period, year_conc = calendar_effect)

# merge
df$year_conc <- NULL
df <- merge(df,df_year, by="year")

# plot
plot(calendar_period, calendar_effect, type="l")
abline(v=aggregate((df$year), list(df$tusk), min)$x, lty ="dashed", col="red")

# Make individual effect
df_age <- df %>% select(tusk = tusk, age = age, age_conc = age_conc)

# create individual intrinsic component which interacts with environmental concentration

# FIRST CHOICE: Exponential decay
# age_year_interaction <- function(x, year_conc){
#   # Exponential individual decay where baseline-factor interacts with yearly concentration in water + slight gamma noise
#   rnorm(length(year_conc), year_conc, 1)*exp(-rnorm(1,0.1,0.01)*x)+rgamma(length(x), 0.1, 1)
# }

# SECOND CHOICE: Constant level
age_year_interaction <- function(x, year_conc){
  # Constant individual level which interacts with yearly concentration in water
  year_conc + rgamma(length(year_conc), year_conc, 1)
}


# create age-effect (confounded by year)
for(tusk in unique(df$tusk)){
  tusk_df <- df_age[df_age$tusk==tusk,]
  byIdx <- which(calendar_period %in% df[df$tusk==tusk,]$year) # calendar period of whale
  baseline <- calendar_effect[byIdx]
  df[df$tusk==tusk,]$age_conc <- age_year_interaction(tusk_df$age, baseline)
}

# plot
ggplot(df, aes(x=age, y=age_conc))+
  geom_line()+
  facet_wrap(.~tusk)

# Fit the model for this element
model <- gamm(age_conc ~ s(age, k=5, bs="cs"), 
              random = list(tusk = ~1), 
              data = df)

# Create prediction data
age_range <- seq(min(df$age, na.rm = TRUE), 
                 max(df$age, na.rm = TRUE), 
                 length.out = 100)

pred_data <- data.frame(age = age_range)

# Generate predictions with confidence intervals
predictions <- predict(model$gam, newdata = pred_data, se.fit = TRUE)

pred_data$fit <- predictions$fit
pred_data$se <- predictions$se.fit
pred_data$lower <- pred_data$fit - 1.96 * pred_data$se
pred_data$upper <- pred_data$fit + 1.96 * pred_data$se

# Create the plot
ggplot(df, aes(x = age, y = age_conc)) +
  geom_line(aes(color = tusk), alpha = 0.5) +
  geom_ribbon(data = pred_data, 
              aes(x = age, y = fit, ymin = lower, ymax = upper),
              alpha = 0.3, fill = "blue", inherit.aes = FALSE) +
  geom_line(data = pred_data, 
            aes(x = age, y = fit), 
            color = "blue", size = 1.2, inherit.aes = FALSE) +
  ylab("Annual concentration") +
  theme_minimal()
```


### SIMULATION EXPERIMENT 3: Linear INCREASE in environmental concentration and exponential decay in individual incorporation


```{r}
# Make calendar effect (linear)
calendar_period <- min(df$year):max(df$year)
sigma = 0.2
calendar_effect <- -190  + 0.1 * calendar_period  + rnorm(length(calendar_period), sd=sigma)

df_year <- data.frame(year = calendar_period, year_conc = calendar_effect)

# merge
df$year_conc <- NULL
df <- merge(df,df_year, by="year")

# plot
plot(calendar_period, calendar_effect, type="l")
abline(v=aggregate((df$year), list(df$tusk), min)$x, lty ="dashed", col="red")

# Make individual effect
df_age <- df %>% select(tusk = tusk, age = age, age_conc = age_conc)

# create individual intrinsic component which interacts with environmental concentration

# FIRST CHOICE: Exponential decay
age_year_interaction <- function(x, year_conc){
  # Exponential individual decay where baseline-factor interacts with yearly concentration in water + slight gamma noise
  rnorm(length(year_conc), year_conc, 1)*exp(-rnorm(1,0.05,0.01)*x)+rgamma(length(x), 0.1, 1)
}

# # SECOND CHOICE: Constant level
# age_year_interaction <- function(x, year_conc){
#   # Constant individual level which interacts with yearly concentration in water
#   year_conc + rgamma(length(year_conc), year_conc, 1)
# }

# create age-effect (confounded by year)
for(tusk in unique(df$tusk)){
  tusk_df <- df_age[df_age$tusk==tusk,]
  byIdx <- which(calendar_period %in% df[df$tusk==tusk,]$year) # calendar period of whale
  baseline <- calendar_effect[byIdx]
  df[df$tusk==tusk,]$age_conc <- age_year_interaction(tusk_df$age, baseline)
}

# plot
ggplot(df, aes(x=age, y=age_conc))+
  geom_line()+
  facet_wrap(.~tusk)

# Fit the model for this element
model <- gamm(age_conc ~ s(age, k=5, bs="cs"), 
              random = list(tusk = ~1), 
              data = df)

# Create prediction data
age_range <- seq(min(df$age, na.rm = TRUE), 
                 max(df$age, na.rm = TRUE), 
                 length.out = 100)

pred_data <- data.frame(age = age_range)

# Generate predictions with confidence intervals
predictions <- predict(model$gam, newdata = pred_data, se.fit = TRUE)

pred_data$fit <- predictions$fit
pred_data$se <- predictions$se.fit
pred_data$lower <- pred_data$fit - 1.96 * pred_data$se
pred_data$upper <- pred_data$fit + 1.96 * pred_data$se

# Create the plot
ggplot(df, aes(x = age, y = age_conc)) +
  geom_line(aes(color = tusk), alpha = 0.5) +
  geom_ribbon(data = pred_data, 
              aes(x = age, y = fit, ymin = lower, ymax = upper),
              alpha = 0.3, fill = "blue", inherit.aes = FALSE) +
  geom_line(data = pred_data, 
            aes(x = age, y = fit), 
            color = "blue", size = 1.2, inherit.aes = FALSE) +
  ylab("Annual concentration") +
  theme_minimal()
```



### SIMULATION EXPERIMENT 4: Periodic change in environment concentration

```{r}
# Make calendar effect (linear)
calendar_period <- min(df$year):max(df$year)
sigma = 0.2
calendar_effect <- 5  + sin(0.5 * calendar_period)  + rnorm(length(calendar_period), sd=sigma)

df_year <- data.frame(year = calendar_period, year_conc = calendar_effect)

# merge
df$year_conc <- NULL
df <- merge(df,df_year, by="year")

# plot
plot(calendar_period, calendar_effect, type="l")
abline(v=aggregate((df$year), list(df$tusk), min)$x, lty ="dashed", col="red")

# Make individual effect
df_age <- df %>% select(tusk = tusk, age = age, age_conc = age_conc)

# create individual intrinsic component which interacts with environmental concentration

# FIRST CHOICE: Exponential decay
# age_year_interaction <- function(x, year_conc){
#   # Exponential individual decay where baseline-factor interacts with yearly concentration in water + slight gamma noise
#   rnorm(length(year_conc), year_conc, 1)*exp(-rnorm(1,0.1,0.01)*x)+rgamma(length(x), 0.1, 1)
# }

# SECOND CHOICE: Constant level
age_year_interaction <- function(x, year_conc){
  # Constant individual level which interacts with yearly concentration in water
  year_conc + rgamma(length(year_conc), year_conc, 1)
}


# create age-effect (confounded by year)
for(tusk in unique(df$tusk)){
  tusk_df <- df_age[df_age$tusk==tusk,]
  byIdx <- which(calendar_period %in% df[df$tusk==tusk,]$year) # calendar period of whale
  baseline <- calendar_effect[byIdx]
  df[df$tusk==tusk,]$age_conc <- age_year_interaction(tusk_df$age, baseline)
}

# plot
ggplot(df, aes(x=age, y=age_conc))+
  geom_line()+
  facet_wrap(.~tusk)

# Fit the model for this element
model <- gamm(age_conc ~ s(age, k=5, bs="cs"), 
              random = list(tusk = ~1), 
              data = df)

# Create prediction data
age_range <- seq(min(df$age, na.rm = TRUE), 
                 max(df$age, na.rm = TRUE), 
                 length.out = 100)

pred_data <- data.frame(age = age_range)

# Generate predictions with confidence intervals
predictions <- predict(model$gam, newdata = pred_data, se.fit = TRUE)

pred_data$fit <- predictions$fit
pred_data$se <- predictions$se.fit
pred_data$lower <- pred_data$fit - 1.96 * pred_data$se
pred_data$upper <- pred_data$fit + 1.96 * pred_data$se

# Create the plot
ggplot(df, aes(x = age, y = age_conc)) +
  geom_line(aes(color = tusk), alpha = 0.5) +
  geom_ribbon(data = pred_data, 
              aes(x = age, y = fit, ymin = lower, ymax = upper),
              alpha = 0.3, fill = "blue", inherit.aes = FALSE) +
  geom_line(data = pred_data, 
            aes(x = age, y = fit), 
            color = "blue", size = 1.2, inherit.aes = FALSE) +
  ylab("Annual concentration") +
  theme_minimal()
```


### SIMULATION EXPERIMENT 5: Sudden change in environment concentration

```{r}
# Make calendar effect (linear)
calendar_period <- min(df$year):max(df$year)
sigma = 0.2
breakpoint <- 2000  # regime shift
calendar_effect <- 5 + ifelse(calendar_period >= breakpoint, 1, 0) + rnorm(length(calendar_period), sd=sigma)

df_year <- data.frame(year = calendar_period, year_conc = calendar_effect)

# merge
df$year_conc <- NULL
df <- merge(df,df_year, by="year")

# plot
plot(calendar_period, calendar_effect, type="l")
abline(v=aggregate((df$year), list(df$tusk), min)$x, lty ="dashed", col="red")

# Make individual effect
df_age <- df %>% select(tusk = tusk, age = age, age_conc = age_conc)

# create individual intrinsic component which interacts with environmental concentration

# FIRST CHOICE: Exponential decay
# age_year_interaction <- function(x, year_conc){
#   # Exponential individual decay where baseline-factor interacts with yearly concentration in water + slight gamma noise
#   rnorm(length(year_conc), year_conc, 1)*exp(-rnorm(1,0.1,0.01)*x)+rgamma(length(x), 0.1, 1)
# }

# SECOND CHOICE: Constant level
age_year_interaction <- function(x, year_conc){
  # Constant individual level which interacts with yearly concentration in water
  year_conc + rgamma(length(year_conc), year_conc, 1)
}

# create age-effect (confounded by year)
for(tusk in unique(df$tusk)){
  tusk_df <- df_age[df_age$tusk==tusk,]
  byIdx <- which(calendar_period %in% df[df$tusk==tusk,]$year) # calendar period of whale
  baseline <- calendar_effect[byIdx]
  df[df$tusk==tusk,]$age_conc <- age_year_interaction(tusk_df$age, baseline)
}

# plot
ggplot(df, aes(x=age, y=age_conc))+
  geom_line()+
  facet_wrap(.~tusk)

# Fit the model for this element
model <- gamm(age_conc ~ s(age, k=5, bs="cs"), 
              random = list(tusk = ~1), 
              data = df)

# Create prediction data
age_range <- seq(min(df$age, na.rm = TRUE), 
                 max(df$age, na.rm = TRUE), 
                 length.out = 100)

pred_data <- data.frame(age = age_range)

# Generate predictions with confidence intervals
predictions <- predict(model$gam, newdata = pred_data, se.fit = TRUE)

pred_data$fit <- predictions$fit
pred_data$se <- predictions$se.fit
pred_data$lower <- pred_data$fit - 1.96 * pred_data$se
pred_data$upper <- pred_data$fit + 1.96 * pred_data$se

# Create the plot
ggplot(df, aes(x = age, y = age_conc)) +
  geom_line(aes(color = tusk), alpha = 0.5) +
  geom_ribbon(data = pred_data, 
              aes(x = age, y = fit, ymin = lower, ymax = upper),
              alpha = 0.3, fill = "blue", inherit.aes = FALSE) +
  geom_line(data = pred_data, 
            aes(x = age, y = fit), 
            color = "blue", size = 1.2, inherit.aes = FALSE) +
  ylab("Annual concentration") +
  theme_minimal()
```